---
title: "Summarize data"
format:
  html:
    code-fold: true
    embed-resources: true
---

## Libraries etc

```{r}
#| warning: false
library(here)

library(data.table)
library(DT)

library(ggplot2)
library(rnaturalearth)
library(dplyr)

library(sf)

read_folder <- here("data/03_grid")
fig_folder <- here("figures/preprocessing")
```


## Load data

```{r}
dat <- readRDS(file.path(read_folder, "french_odonate_data.rds"))

dim(dat)
str(dat)
```


Variable| Description
:-----|:----------
scientificName| latin name of the observed taxa
family| family name of the observed taxa
occurrenceID| ID of the occurence event (unique ID of the row)
eventDate| date of the observation
recordedBy| name of the observer
decimalLongitude| longitude in decimal degree (WGS84)
decimalLatitude| latitude in decimal degree (WGS84)
municipality| municipality of the observation
verbatimName| full name of the species, including reference
genus| genus of the taxa   
taxonRank| rank of the taxa
taxonID| ID of the taxa (from GBIF)              
id_1000| ID of the grid cell at 1km resolution
id_5000| ID of the grid cell at 5km resolution
id_10000| ID of the grid cell at 10km resolution
source| database of origin (steli or atlas)                
x_ETRS89_LAEA| longitude in meter following the ETRS89 coordinate reference system and the Lambert Azimuthal Equal-Area projection
y_ETRS89_LAEA| latitude in meter following the ETRS89 coordinate reference system and the Lambert Azimuthal Equal-Area projection
region| administrative region of the observation
departement| administrative department of the observation
elevation_m| elevation (in m), from [Copernicus GLO-30 dataset](https://dataspace.copernicus.eu/explore-data/data-collections/copernicus-contributing-missions/collections-description/COP-DEM)
GEnS_v3_bioclim| bioclimatic region, from [Metzger et al. 2013](https://doi.org/10.1111/geb.12022)
popdensity_hab_per_km2| population density in 2006 from [EEA Geostat data](ttps://www.eea.europa.eu/en/datahub/datahubitem-view/5884f314-84a9-4f53-a745-d94d7a53e8b1?activeAccordion=1083668%2C761)
CLC2018_landcover| [Corine land cover class](https://land.copernicus.eu/en/products/corine-land-cover/clc2018) of 2018 with spatial resolution of 100m

## Summarize data

### Tables

```{r}
# Number of observations by program
dat[, .N, by = source] |> 
  datatable()

# Number of species
nsp <- dat[, .N, by = scientificName][order(-N), ]

nsp |> datatable()
```


## Species barplot

```{r, fig.height=6, fig.width=8}
ggplot(nsp) +
  geom_col(aes(x = N, 
               y = reorder(scientificName, N)), fill = "darkorange") +
  scale_x_continuous(expand = c(0,0)) +
  xlab("Number of occurrences") +
  theme_minimal(base_size = 16) +
  theme(axis.title.y = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(face = "italic",
                                   size = 6))

ggsave(file.path(fig_folder, "spp_barplot.png"),
       width = 8, height = 6, dpi = 300)
```


### Data evolution

```{r, fig.width=6, fig.height=3}
datN <- dat[, .N, by = list(source, year(dat$eventDate))]

ggplot(datN) +
  geom_line(aes(x = year, y = N)) +
  # geom_point(aes(x = year, y = N)) +
  theme_minimal() +
  ylab("Occurrences") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  # scale_x_continuous(breaks = c(seq(1700, 1900, by = 100),
  #                               seq(1990, 2024, by = 10))) +
  # # scale_x_continuous(breaks = seq(2010, 2025, by = 5)) +
  facet_wrap(facets = vars(source), 
             scales = "free") +
  theme(axis.title.x = element_blank())

ggsave(file.path(fig_folder, "dat_evolution.png"),
       width = 6, height = 3, dpi = 300)
```

### Map

```{r}
dat_sf <- st_as_sf(dat,
                   coords = c("x_ETRS89_LAEA", "y_ETRS89_LAEA"))
st_crs(dat_sf) <- 3035

# Data bounding box
bbox <- st_bbox(dat_sf)
```

Plot raw data points:

```{r}
# Get French map
fr <- ne_countries(scale = 50, 
                   country = "France")
fr_3035 <- st_transform(fr, 3035)
```

```{r}
#| eval: false
ggplot() +
  geom_sf(data = fr_3035) +
  xlim(bbox$xmin, bbox$xmax) +
  ylim(bbox$ymin, bbox$ymax) +
  geom_sf(data = dat_sf,
          aes(col = source), size = 0.1) +
  facet_wrap(facets = vars(source))
```

Plot data on grid:

```{r}
# Get grid
grid <- readRDS(file.path(read_folder, 
                          "grid_scale_10000.rds"))
```

```{r}
# Sumarize with grid
nobs <- st_drop_geometry(dat) |>
  group_by(id_10000, source) |>
  summarize(nobs = n(), .groups = "drop")

nobs_grid <- grid |>
  right_join(nobs, by = c("id" = "id_10000"))
```

```{r}
# Plot data
ggplot() +
  geom_sf(data = fr_3035) +
  geom_sf(data = nobs_grid, aes(fill = nobs),
          color = "transparent") +
  facet_grid(cols = vars(source)) +
  xlim(bbox$xmin, bbox$xmax) +
  ylim(bbox$ymin, bbox$ymax) +
  scale_fill_viridis_c(na.value = "lightgrey",
                       trans = "log",
                       breaks = c(1, 20, 400, 8000),
                       name = "Occurrences") +
  theme_void(base_size = 30) + 
  theme(legend.position = "left",
        legend.key.height = unit(1, "cm"))

ggsave(file.path(fig_folder, "map.png"),
       width = 8, height = 6, dpi = 300)
```


